---
title: "Performing PCA and FA"
author: "Rob Kabacoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performing PCA and FA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **qacDR** package provides functions for principal components and common factor analysis. Based on the William Revelle's comprehensive \link[psych]{https://cran.r-project.org/web/packages/psych/index.html} package, the \code{factorAnalysis} package provides simplified input, intuitive output, and easily interpretable graphs, making these techniques more accessible to data analysts new to these forms of analysis.

## Principal Components Analysis

Let's perform a principal components analysis on some ratings data. First we'll import the data into R.

```{r}
# input data
ratings <- read.csv("https://www.promptcloud.com/wp-content/uploads/2017/02/EFA.csv")
head(ratings)
```

Next, well create a scree plot of the data.

```{r fig.width=6, fig.height=6}
library(qacDR)

# scree plot
scree_plot(ratings)
```

The scree plot suggests four components. In the next step, we'll extract four principal components and rotate them using a *varimax* rotation.

```{r fig.width=6, fig.height=6}
# extract 2 principal components
fit.pca <- PCA(ratings, nfactor=4, rotate="varimax")
```

The four components account for 55% of the variance in the original data. Next, we'll plot the pattern matrix as both a table and a bar chart.

```{r fig.width=6, fig.height=6}
# plot factor pattern as bar chart
plot(fit.pca, sort=TRUE)

# plot factor pattern as a table
plot(fit.pca, sort=TRUE, type="table")
```

Finally, we'll add the component scores to the original data.

```{r}
# save component scores
mydata <- score(ratings, fit.pca)
head(mydata)
```

## Factor Analysis

In this section, the ratings data are re-analyzed using a *maximum likelihood* factor analysis.

Again, we'll start with a scree plot. This time we'll run 200 simulations. A random number seed is set to assure reproducability.

```{r fig.height=6, fig.width=6, message=FALSE, warning=FALSE, message=FALSE}
library(qacDR)

# scree plot
set.seed(1234)
scree_plot(ratings, method = "ml", n.iter=200)
```

Again, the scree plot suggests two factors. In the next step, we'll extract four factors and rotate them using a *promax* rotation.

```{r fig.width=6, fig.height=6}
# extract 4 factors
fit.fa <- FA(ratings, nfactor=4, rotate="promax", fm="ml")
```

The printout includes the factor pattern, the factor structure, and the factor intercorrelations. The four factors account for 40% of the variance in the original data. The largest factor intercorrelation is between factor 1 and factor 3 (0.37) .

Next, we'll plot the pattern matrix as both a table and a bar chart.

```{r fig.width=6, fig.height=6}
# plot factor pattern as table
plot(fit.fa, sort=TRUE)

# plot factor pattern as bar chart
plot(fit.fa, sort=TRUE, type="table")
```

Finally, we'll add the factor scores to the original data.

```{r}
# save component scores
mydata <- score(ratings, fit.fa)
head(mydata)
```
